---
title: "Final Project"
author: "Liz Austin, Casey Gibson, and Zhengfan Wang"
date: "November 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Lyme Disease CDC Data
Dataset detailing cases of Lyme Disease per 100,000 population in each state for the years 2006-2016

Data source:
https://www.cdc.gov/lyme/stats/index.html


Clustering by region reference:
https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf

#Group Analysis Proposal
1. Simple Poisson Model 
2. Model with state as random effect
3. Model with state as a covariate 
4. Model with region as a random effect
5. Model with region as a covariate

*Compare*

6. Select "baseline" model to simulate data from 

#Individual Analysis Proposal

Ideas in general:

1. Gamma-poisson mixture model (neg-binom)
2. Autoregressive model on $\lambda$'s 
3. Consider adding a penalized spline
4. Priors on $\beta$'s/bayesian land
5. Simulation

###Liz
Poisson Models for Overdispersion with simulation:

  + Poisson GLM with $\hat\phi$ adjustment 
  
  + Poisson GLMM with normal random intercept
  
  + Gamma-Poisson mixture model
  
  + GEE marginal Poisson
  
###Casey
Bayesian + Autoregressive Model

###Zhengfan
Clustering + Simulation

```{r, echo=FALSE}
lyme=read.csv("lymedata.csv",header=T)

#Let 1 = Northeast
#Let 2 = Midwest
#Let 3 = South
#Let 4 = West
lyme$Region<- c(3,4,4,3,4,4,1,3,3,3,3,4,4,2,2,2,2,3,3,1,3,1,2,2,3,2,4,2,4,1,1,4,1,3,2,2,3,4,1,1,3,2,3,3,4,1,3,4,3,2,4)
lyme<- lyme[-9,]

lyme1=reshape(lyme, direction="long", varying=list(names(lyme)[2:11]), v.names="count", 
        idvar=c("state"), timevar="year", times=2006:2015)
lyme1<- lyme1[,-5]
lyme1$count<- as.numeric(lyme1$count)

mod1<- glm(count~year,data=lyme1,family=poisson)

mod2<- glm(count~year+State,data=lyme1,family=poisson) 

mod3<- glm(count~year+Region,data=lyme1,family=poisson) 

anova(mod2,mod3)
505-456
312060-20576
pchisq(291484, df=49, lower.tail=FALSE)

lyme1$count<- as.integer(lyme1$count)
lyme1$Region<- as.integer(lyme1$Region)

#mod4<- glmm(count~year, random=list(~ 0 + Region), varcomps.names = c("1","2","3","4"), data=lyme1, family.glmm=poisson.glmm)
```

#Dataset sample 

```{r}
lyme1[1:10,]
lyme1[51:60,]
```

```{r}
#simulation
mod3
n<- 1
alpha<- -6.325388
beta_1<- 0.007653 
beta_2<- -1.461143
set.seed(102)
y <- c()
for (i in 1:500){
  y <- c(y,rpois(1,lambda = exp(alpha+beta_1*lyme1$year[i] + beta_2*lyme1$Region[i])))
}

simulated_data <- data.frame(incidence=y,year = lyme1$year, region=lyme1$Region)
View(simulated_data)

require(rbiips)
library(MCMCpack)

n_part = 10000 # Number of particles
variables = c('x','y') # Variables to be monitored
mn_type = 'fs'; rs_type = 'stratified'; rs_thres = 0.5 # Optional parameters



out_smc = biips_smc_samples(model, variables, n_part,
                            type=mn_type, rs_type=rs_type, rs_thres=rs_thres)
diag_smc = biips_diagnosis(out_smc)
summ_smc = biips_summary(out_smc, probs=c(.025, .975))


x_f_mean = summ_smc$x$f$mean
x_f_quant = summ_smc$x$f$quant


```
